\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/patients.tex
All text

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}~\cite{Peters2022}.

\todo{visual bci}

@@

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Patients with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work an literature to
people with \ac{sspi} and various degrees motor impairment in a visuall oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of individuals with
    \ac{sspgi}, when operating a visual \ac{bci},
  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
    group,
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, participants must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Patient PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abcess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

@@

During the recording session, participants were positioned in their wheelchair in front of a table.
Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
Inc., Taiwan) placed at a 60cm distance.
A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking throughout was performed using the Tobii~\todo{model, make}
portable eye tracker placed at the bottom of the laptop screen.

@@

\subsection{BCI decoding}

\todo{shortly describe WCBLE}
\todo{shortly describe tLDA}
\todo{shortly describe xdawncov+ts+LDA}

@@

It was technically impossible to register gaze position with the Tobii
\todo{model} for participants PC2 and PC3 since they both had one eye that was
occluded respectively by the prism glass and the eyelid.
Both participants reported they could not fixate some of the
stimuli.

\subsection{\Ac{bci} decoding performance}

Figure~\ref{fig:patients/decode} shows single-trial
\ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.
Scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \ac{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    settign.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}
\todo{figure margin}
\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}

@@

\subsection{Cross-condition calibration}
\label{sec:patients/outcomes/cross}

\todo{input and describe graph}

@@

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspgi}.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspgi}.

@@

Individuals with all but the most severe gaze impairments will probably retain
some degree of gaze direction in visual \ac{bci} operation, which can
drastically boost performance.
Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with his
overt \ac{vsa} performance, although eye tracking showed that he relied mostly
on over \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
gaze was uncued.
This is also supported by our results on cross condition calibration\todo{how}

\todo{discuss that training in overt always increases covert/free, focus on
increased P3 amplitude and relying on visual components, also propose this
strategy if possible}

@@

Despite results that prompt reflection on gaze-independent \ac{bci} approaches,
there are some limitations to the presented results that need to be
addressed in current and future work.

First and foremost, this study works with a limited sample size which
does not represent the full spectrum of individuals with \ac{sspi} and
\ac{sspgi}, and their specific symptoms and skills?
Individuals with \ac{fa} met the inclusion criteria, but they are usually not
considered as one of the typical interest groups of \ac{bci} communication assistive
technology, partly due to the rarity, and partly due to the disease
progression.
It would be most interesting to verify these results with individuals with
\ac{lis} and no eye movement capability at all.

@@

In this study, user comfort in the different conditions was not objectively
measured.
Rather, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
In order to properly contextualize performance results, they should be coupled
with a measure of user comfort and percieved effort, like the NASA Task Load
Index~\cite{Hart2006}.
Performance might after all be traded off for user comfort.
Eye motor disability can also have been assessed more objectively, using \todo{

The Coma Recovery Scale
Revised (CRS-R) [73] provides standardized procedures for admin-
istration and response criteria to more uniformly describe the vis-
ual fixation or pursuit skills of a user with SSPI. The Northeastern
State University College of Optometry (NSUCO) Oculomotor Test
[74] provides specific procedural techniques as well as scoring cri-
teria, which allows for uniform description of the accuracy and
ability of pursuit and saccade skills for adults and children.
}

@@

To summarize, we present our three main findings.
\begin{enumerate*}
  \item We have confirmed that some individuals with \ac{sspi} and gaze
    impairment find it more comfortable to operate a visual gaze-independent
    \ac{bci} relying on covert \acf{vsa}.
  \item We have shown that covert \acf{vsa} with central gaze fixation can be
    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
    but this does not hold true if central gaze fixation is not imposed.
  \item We have shown that covert \ac{vsa}
    decoding can alternatively be improved by
    using a calibration session with attempted overt \ac{vsa}.
\end{enumerate*} \todo{with or without imposed gaze fixation}

@@

\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
\todo{BCI principles and practice chapter 19}
\todo{https://www.nature.com/articles/s44222-024-00239-5}

>>>>

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}~\cite{Peters2022}.
Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
offer several advantages in this context.
They can work with non-invasive recording technology and can use rapid
stimulation.
This makes them well-suited for real-time tasks communication tasks.

@@

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Individuals with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work an literature to
people with \ac{sspi} and various degrees motor impairment in a visuall oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of individuals with
    \ac{sspgi}, when operating a visual \ac{bci},
  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
    group,
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Recruitment}
Participants were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
\todo{include years since diagnosis}

@@

Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Participant PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abcess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

@@

During the recording session, participants were positioned in their wheelchair in front of a table.
Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
Inc., Taiwan) placed at a 60cm distance.
A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking throughout was performed using the Tobii X2-30 Compact (Tobii
Technology AB, Sweden) portable eye tracker placed at the bottom of the laptop screen.

@@

\subsection{\Acs{bci} decoding}

We evaluate the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
and \ac{tlda}~\cite{Sosulski2022}.
classifiers, and the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
For \ac{wcble}, a region of interest from 0ms to 800ms relative to stimulus
onset was used while the epoch was cropped to -100ms to 900ms, for the other
decoders the epoch was cropped between 0ms and 800ms which resulted in maximal
performance.
Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

refer to~\autoref{sec:covert-align/methods/decoders}.

@@

It was technically impossible to register gaze position with the Tobii X2-30
Compact for participants PC2 and PC3 since they both had one eye that was
occluded respectively by the prism glass and the eyelid.
Both participants reported they could not fixate some of the
stimuli.

\subsection{\Acs{bci} decoding performance}

Figure~\ref{fig:patients/decode} shows single-trial
\ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \acs{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    settign.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}
\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}

@@

\subsection{Cross-condition calibration}
\label{sec:patients/outcomes/cross}
As an alternative approach to picking the most suitable decoder, we chose
\ac{tlda} as a given decoder and verified whether performance can be improved
if \ac{bci} users with gaze impairment perform the calibration session relying
maximally on their residual gaze control.

\begin{figure}[t]
  \input{figures/patients/fig_cross.pgf}
  \caption[Cross-setting calibration and decoding performance.]{
    Decoding performance when calibrating the \ac{tlda} decoder in a given \ac{vsa}
    setting, and evaluating it in another, reported as
    single-trial \ac{rocauc}.
    For participant PA1, free \ac{vsa} performance could be improved when
    calibrating with overt gaze fixation.
    For participants PA1, PB2 and PC3, this was the case when operating in the
    covert \ac{vsa} setting with central gaze fixation and calibrating with
    overt gaze fixation.
  }
  \label{fig:patients/cross}
\end{figure}

\autoref{fig:patients/cross} shows that, on average, covert~\ac{vsa} decoding
can be improved when training in overt \ac{vsa}.
This was especially the case for participants PA1, PB2 and PC3.
Note that, according to eye tracking data, patients PB1, PB4 and PC4 did not
always perform the cued central gaze fixation in the covert \ac{vsa} setting,
which might affect the results.

@@

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspgi}.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024}
presented in~\autoref{sec:covert-align} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspgi}.

@@

Individuals with all but the most severe gaze impairments will probably retain
some degree of gaze direction in visual \ac{bci} operation, which can
drastically boost performance.
Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with his
overt \ac{vsa} performance, although eye tracking showed that he relied mostly
on over \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
gaze was uncued.
This is also supported by our results on cross condition calibration presented
in~\autoref{sec:patients/outcomes/cross}, which show that leveraging residual
eye motor control to fixate targets during the calibration phase can improve
performance in some settings.
This is likely due to the increased P3 component amplitude in overt \ac{vsa},
which improves the discriminative power of a classifier trained on this data.
Cueing this overt gaze fixation only during the calibration phase leaves user
free to operate in the manner that is most comfortable for them in  the
operation phase.
Early VEPs in the training data could also contribute in those cases where the participant was not
able to perform covert \ac{vsa} with central gaze fixation.

@@

Despite results that prompt interesting reflections on gaze-independent \ac{bci} approaches,
there are some limitations to the presented results that need to be
addressed in current and future work.

First and foremost, this study works with a limited sample sizekkjk which
does not represent the full spectrum of individuals with \ac{sspi} and
\ac{sspgi}, and their specific symptoms and skills?
Individuals with \ac{fa} met the inclusion criteria, but they are usually not
considered as one of the typical interest groups of \ac{bci} communication assistive
technology, partly due to the rarity, and partly due to the disease
progression.
It would be most interesting to verify these results with individuals with
\ac{lis} and no eye movement capability at all.

@@

In this study, user comfort in the different conditions was not objectively
measured.
Rather, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
In order to properly contextualize performance results, they should be coupled
with a measure of user comfort and percieved effort, like the NASA Task Load
Index~\cite{Hart2006}.
Performance might after all be traded off for user comfort.
Eye motor disability can also have been assessed more
objectively~\cite{FriedOken2020}, using e.g.\
the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
oculomotor~\cite{Maples1992}.

@@

To summarize, we present our three main findings.
\begin{enumerate*}
  \item We have confirmed that some individuals with \ac{sspi} and gaze
    impairment find it more comfortable to operate a visual gaze-independent
    \ac{bci} relying on covert \acf{vsa}.
  \item We have shown that covert \acf{vsa} with central gaze fixation can be
    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
    but this does not hold true if central gaze fixation is not imposed.
  \item We have shown that decoding of covert \ac{vsa}
    can alternatively be improved by
    using a calibration session with attempted overt \ac{vsa}.
\end{enumerate*}

@@
